---
title: "Tutorial on estimating zebra mussel density"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tibble.width = Inf)
```


This tutorial assumes that you have collected data following one of the survey procedures documented in the MAISRC video, ["Monitoring Zebra Mussels in Lakes"](https://vimeo.com/317738991/9189d0dcc0). Here, we describe how to analyze data collected using dependent double-observer surveys with and without distance sampling. These designs are sometimes refered to as "removal designs" (rather than dependent double-observer designs). We will demonstrate these techniques using data collected on Lake Burgan, Minnesota in the summer of 2018.  Both of these survey designs allow you to estimate the probability of detecting  mussels, and thus,  obtain estimates of density adjusted for imperfect detection. Here we show how to use design-based estimators of density. All data and files are available for download [here](#links).

# Estimating density and quantifying uncertainty

Our goal is to estimate zebra mussel density in a lake and to quantify uncertainty associated with that estimate. To do this, we need the counts in each transect, denoted as $x_i$ for the $i$th transect, the transect length, $l_i$, and the width, $w$, to determine the surveyed area, $a_i=w l_i$.  The degree of uncertainty in our density estimate will depend on the variance of the counts among the different transects, $x_i$, and the variance in the estimated detection probability, $\mathrm{var}(\hat{P})$, where $\hat{P}$ denotes the estimated probability of detection of a zebra mussel by either observer. Here are the formulae we used to estimate density and its variance:

\begin{align}
  \hat{D} &= \frac{\displaystyle\sum_i^n  x_i}{\hat{P} \sum_i^n  w l_i}\label{eq:hatD}\\
  \mathrm{var}(\hat{D}) &= \hat{D}^2 \left(  \frac{\mathrm{var}(\sum_i^n  x_i)}{(\sum_i^n  x_i)^2} + \frac{\mathrm{var}(\hat{P})}{\hat{P}^2} \right). \label{eq:varD}
\end{align}

Our estimate of $\mathrm{var}(\hat{P})$ will depend on the survey design.

# Dependent Double-observer (no distance data)

To conduct the dependent double-observer  survey, 2 divers  survey 30-meter x 1-meter wide areas deliniated using 2 lead lines laid in parallel. The primary diver swims first, marking all the mussels she/he sees within the belt. The secondary diver follows looking for mussels that the first observer missed. We rotated the role of primary and secondary divers at each transect. 

## The data

Data should be entered in a manner that ensures reliability. For this dataset, we entered data using Google Forms. This software allowed us to check that no impossible values were entered (for example, transect lengths were numbers between 1 and 30 due to our design). We entered data of each detection event (called EncountersDoubleTutorial.xlsx) and a sheet that described each transect (TransectsDoubleTutorial.xlsx). The unique variable "Transect number" links these sheets. 
Read in the encounter data:
```{r}
suppressPackageStartupMessages(library(tidyverse)) #loads the tidyverse quietly
library(readxl)
double.dat      <- read_xlsx(path="EncountersDoubleTutorial.xlsx", sheet=1)
```
And read in the transect data:
```{r}
transect.dat    <- read_xlsx(path="TransectsDoubleTutorial.xlsx", sheet=1)
transect.dat    <- transect.dat %>% gather(key = observer, value = name, "primary", "secondary") 
```

Now we need to format the data for the analysis. We will be using the `removal` function in the R fisheries stock analysis (`RSA`) package to calculate the estimates. This function will estimate the detection probability and variance in counts for us. We can use this information to determine the variance in density. We have written the `create.removal` function to properly format the data and have included it in the file [`HelperFuncs.R`](HelperFuncs.R), which we source here (making the function available to us in the current R session). Tutorials on the estimator used here are available at the [FishR](http://derekogle.com/fishR/examples/oldFishRVignettes/Depletion.pdf) site. 

```{r, warning=F}
source('HelperFuncs.R')

#returns data in a form that can be used by removal in FSA
removal.list <- create.removal(double.dat, transect.dat) 
```


## The analysis

Here we apply the function `removal` to each transect seperately, then combine the estimates by weighting by transect length. This function calculates the variance for us so we don't need to apply the variance formula.
```{r, warning=F, messages=F}
library(FSA)
#get the length of each transect
length.vec  <- transect.dat %>% group_by(`Transect number`) %>% summarize(length = first(length)) 

remove.ests <- removal(removal.list, just.ests=TRUE)

# now sum the results over all transects to estimate
# population size at the transect level 
Nhat       <- remove.ests[1]
Nhat.SE    <- remove.ests[2]#remove.ests[1]sqrt(sum(data.frame(remove.out)[2,]^2, na.rm = TRUE))

# estimate of density in the transect
Dhat       <- Nhat/sum(length.vec$length)
Dhat.SE    <- Nhat.SE/sum(length.vec$length)

# estimate of detection probability in the transect
bp.hat      <- remove.ests[5]
bpSE.phat   <- remove.ests[6]

```

We get an estimate of density $\hat{D}=$ `r round(Dhat,2)` mussels per square meter with a standard error of `r round(Dhat.SE,2)`. We estimate the detection to be $\hat{P}=$ `r round(bp.hat,2)` with a standard error of `r round(bpSE.phat,2)`,

# Dependent Double-observer Survey with Distance Data

The dependent double-observer survey with distance data is  similar to the dependent double-observer survey described above. However, in this case, both divers swim along a single transect line (lead line).  Further, whenever a diver detects a mussel or cluster of mussels, the diver also measures the perpendicular distance from the detected mussel(s) to the transect line.  The secondary diver again follows the primary diver, looking for mussels that were missed by the primary diver. These distance to detection measurements are then used to model how detection probabilities decline with distance from the transect line. This distance model is then used to correct counts for imperfect detection. The [distancesampling.org](http://distancesampling.org/R/index.html) website provides tutorials on using the R package described here, as well as several other packages that can be used to do distance sampling with either single- or double-observer designs.

## The data

Here we enter each dataset. Read in the encounter data:
```{r}
suppressPackageStartupMessages(library(tidyverse)) #loads the tidyverse quietly
distance.dat      <- read_xlsx(path="EncountersDistanceTutorial.xlsx", sheet=1)
```
And read in the transect data:
```{r}
transect.dat    <- read_xlsx(path="TransectsDistanceTutorial.xlsx", sheet=1)
transect.dat    <- transect.dat[order(transect.dat$"Transect number"),]
```

It is useful to construct some initial diagnostic plots to ensure  data were entered correctly and assumptions of distance sampling are met (e.g., detection declines monotonically with distance from the transect line). 
```{r pressure, fig.cap="Diagnositic plot of the detection distance, the distance of each detection from the transect line.", fig.asp=0.7}
hist(distance.dat$distance, col="black", border="white",
     xlab='Distance of detection from transect (cm)', main="")
```

Here we see a pattern that is consistent with our expectation - i.e., the number of detections  declines with distance from the transect line. Now we'll apply the distance survey analysis.

## The analysis

Here we use a helper function to format the data appropriately for the R package `MRDS`. 

```{r, warning=F}
library(mrds)
source('HelperFuncs.R')
distance.dat      <- distance.dat[order(distance.dat$`Transect #`),]

distance.dat      <- distance.dat %>% mutate(detected = rep(1, dim(distance.dat)[1]), object=1:dim(distance.dat)[1]) 
distance.rem.dat  <- create.removal.Observer(transect.dat=transect.dat, obs.dat=distance.dat) 

```

Based on the diagnostic figure of detection distances we chose  to use the hazard rate model (`dsmodel=~cds(key="hr")`) to capture the drop-off of detection with distance. Alternative distance functions in the `mrds` package are the half-normal (`hn`), gamma function (`gamma`), and uniform (`unif`).
```{r, warning=F}
distance.model  <- ddf(method="rem", dsmodel=~cds(key="hr"), mrmodel=~glm(formula=~1), data=distance.rem.dat, meta.data=list(width=100))

print(summary(distance.model))
```

The average probability of detecting a mussel by one of the observers is $\hat{P}=0.58$ with a standard error of $\sigma_\hat{P}=0.06$. We can also plot the  distance detection function. There are several possible diagnostic plots to look at. Below, we overlay the estimated detection model on the histogram of detection distances made by the two observers. 
```{r, warning=F}
plot(distance.model, which=2, showpoints=FALSE, lwd=2)
```

The variance in the total counts can be estimated using the design estimator, $\mathrm{var}(X) = \left(L \sum_{i}^n l_i (x_i/l_i - X/L)^2 \right)/(n-1)$. Below we calculate the estimated density and it's uncertainty by by using the output from the distance model.

```{r}
area    <- transect.dat$`Transect length (if transect survey)` #length of transects
counts  <- table(distance.dat$`Transect #`) #count in each transect
n       <- length(area) #total number of transects
w       <- 2 #width of transects
Phat    <- 0.58 #from distance model
VarP    <- 0.06^2 #from calculation above

#get density esimate and its standard error.
Dhat    <- sum(counts)/(w*sum(area)*Phat)
VarX    <- sum(area)*sum(area*(counts/area - sum(counts)/sum(area))^2)/(n-1)
  
VarD <- Dhat^2*(VarX/sum(counts)^2 + VarP/Phat^2)
```

We get an estimate for $\hat{D}=$ `r round(Dhat, 2)` with a standard error of $\sqrt{\mathrm{var}({\hat{D}})}=$ `r round(sqrt(VarD), 2)`.

## Resources

Book: [Buckland et al. (2015) Distance sampling: methods and applications.](https://www.springer.com/us/book/9783319192185)

[distancesampling.org](http://distancesampling.org/R/index.html). Tutorials on distance sampling in R.

[FishR](http://derekogle.com/fishR/examples/). Tutorial on depeletion sampling in R.

### Files used in this tutorial {#links}

[Distance survey detections](https://github.com/troutinthemilk/Zeebs-Season2/raw/master/Tutorial/EncountersDistanceTutorial.xlsx)

[Distance survey transects](https://github.com/troutinthemilk/Zeebs-Season2/raw/master/Tutorial/TransectsDistanceTutorial.xlsx)

[Double-observer survey detections](https://github.com/troutinthemilk/Zeebs-Season2/raw/master/Tutorial/EncountersDoubleTutorial.xlsx)

[Double-observer survey transects](https://github.com/troutinthemilk/Zeebs-Season2/raw/master/Tutorial/TransectsDoubleTutorial.xlsx)

[HelperFuncs.R](https://github.com/troutinthemilk/Zeebs-Season2/raw/master/Tutorial/HelperFuncs.R)

[Tutorial markdown file](https://github.com/troutinthemilk/Zeebs-Season2/raw/master/Tutorial/Tutorial.Rmd)



